### 2020-11-08 serial.tex
submitted
no competing interests
format & supp

\usepackage{natbib}
\bibliographystyle{chicago}
\date{\today}

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic, such as the current pandemic of coronavirus disease 2019 (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case.
The value in a fully susceptible population --- the ``basic'' reproduction number \Ro --- allows us to predict the extent to which an infection will spread in the population, and the amount of intervention necessary to eliminate it in simple cases \citep{anderson1991infectious}.
Since the reproduction number represents an average \citep{diekmann1990definition, anderson1991infectious}, it fails to capture heterogeneity among individuals or across space.
The reproduction number also fails to provide any information about the time scale of disease transmission.

Estimating the reproduction number \RR is often challenging.
Direct estimates based on observed infections will typically be biased down when some infections cannot be observed.
A common method of estimating \RR near the beginning of an epidemic is based on the population-level exponential growth rate $r$, which can often be estimated robustly from case reports \citep{mills2004transmissibility, ma2014estimating}.
The growth rate $r$ and the reproduction number \RR are linked by the generation-interval distribution \cite{wallinga2007generation}, where the generation interval is defined as the time between when an individual (infector) is infected and when that individual infects another person (infectee) \citep{svensson2007note}.

Since generation intervals measure time between infection events, which can be difficult to observe in practice, generation intervals are often replaced with serial intervals.
The serial interval is defined as the time between when an infector and an infectee develop symptoms \citep{svensson2007note}.
While generation and serial intervals both measure the time scale of disease transmission, they measure fundamentally different quantities. 
In particular, previous studies have noted that, in many contexts, serial intervals are expected to have larger variances than generation intervals but have the same mean in many contexts \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence}.
Serial intervals can in some cases even take negative values in the presence of presymptomatic transmission \citep{he2020temporal}, whereas generation intervals must be positive.

Although these distributions were clearly and distinctly defined over a decade ago \citep{svensson2007note}, 
the need for a better conceptual and theoretical framework for understanding their differences is becoming clearer as the COVID-19 pandemic unfolds.
Researchers continue to base inferences about COVID-19 on both generation and serial intervals without clearly distinguishing between them (e.g., \cite{tempvar,du2020serial,he2020temporal,wu2020nowcasting,zhao2020serial}), and, in some cases, explicitly conflate the definitions of the two intervals (e.g., \cite{anderson2020will,hellewell2020feasibility}).
This confusion is apparent even in standard software for estimating \RR, such as \texttt{EpiEstim}, in which the serial-interval distribution is used to infer time-dependent \RR \citep{thompson2019improved}.
These studies are examples of many---indeed, it is a common practice to use the serial and generation intervals interchangeably.

One source of confusion arises from an apparent discrepancy between the generation-interval and serial-interval viewpoints.
While the epidemic is growing exponentially, the spread of infection can be characterized as a \emph{renewal process} based on previous incidence of infection, the associated generation-interval distribution, and the average infectiousness of an infected individual.
It is well established that this renewal formulation allows us to link the exponential growth rate of an epidemic $r$ with its reproduction number \RR using the generation-interval distribution \citep{wallinga2007generation}.
However, the serial-interval distribution also describes a renewal process --- in this case, the creation of a new \emph{symptomatic} case based on a symptomatic case in the previous generation.
Since both renewal processes, based on either generation- or serial-interval distributions, describe the same underlying exponentially growing system, both should provide the same correct link between the reproduction number \RR and the epidemic growth rate $r$.

In contexts where the serial- and generation-interval distributions differ, current theory has no explanation for how two different distributions could provide identical estimates of \RR from $r$. 
In fact, recent theory suggest that using the serial-interval can underestimate the reproduction number \citep{britton2019estimation, ganyani2020estimating}.
However, these studies rely on \emph{intrinsic} distributions of incubation periods and generation intervals that neglect the population-level dynamics of disease spread.

Here we show that, by correctly defining and calculating the ``forward'' serial-interval distribution (i.e., a distribution of serial intervals from a cohort of infectors that developed symptoms at the same time) that connects symptom onset dates, we can resolve this discrepancy.
These forward intervals are different from the ``intrinsic'' serial intervals that previous studies have relied on \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence, britton2019estimation}.
During an ongoing epidemic, all observed epidemiological delays (e.g., incubation period) between primary (e.g., infection) and secondary (e.g., symptom onset) events are subject to backward biases: when the incidence of primary events is increasing (or decreasing), we are more likely to observe shorter (respectively longer) intervals.
In particular, when we consider forward serial-interval distributions, the incubation periods of the infectors are subject to backward biases because we have to look backward in time from their symptom onset to infection.
Therefore, the realized incubation period distributions of the infector and the infectee can differ dynamically, even if the intrinsic analogues of the same distributions are expected to be equivalent.

@@

At the individual level, we can define the time distribution between a primary and a secondary event that we expect to observe for a single infected individual by averaging across individual characteristics --- we refer to this distribution as the \emph{intrinsic distribution}.
For example, the intrinsic incubation period distribution describes the expected time distribution from infection to symptom onset of an infected individual.
Likewise, the intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by an infected individual.
However, the intrinsic time distributions are not always equivalent to the corresponding realized time distributions at the population level (i.e., the distribution of time between actual primary and secondary events that occur during an epidemic; see \fref{diagram}).
For example, an infectious contact results in infection only if the contacted individual is susceptible (and has not already been infected) ---
this is one mechanism that causes realized generation intervals (time between actual infection events) to differ from from the intrinsic generation intervals (time between infection and infectious contacts) \citep{park2020inferring}.
In this example, the difference between intrinsic and realized time distributions can be attributed to the fact that the fraction of susceptible individuals is itself dynamic.

@@

Several different mechanisms drive the changes in forward and backward delay distributions over time.
Typically, within-individual forward delay distributions are not directly affected by epidemic dynamics.
Some realized forward distributions, like incubation period distributions, are equivalent to their intrinsic distributions and remain invariant at the time scale of an outbreak.
Other realized distributions, like the distribution of time from symptom onset to testing, may change over the course of an epidemic due to changes in public-health policies or individual behavior.
Between-individual forward delay distributions, such as generation- or serial-interval distributions, depend on epidemic dynamics.
For example, forward generation intervals often become shorter as an epidemic progresses due to the dynamical process of susceptible depletion, as well as due to other factors like behavioral change or interventions \citep{kenah2008generation, nishiura2010time, champredon2015intrinsic}: if it is harder to infect later in the course of infection, then proportionally more intervals will be short.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when their infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau_{\rm s}$ in the form (\fref{diagram}A):
\begin{equation}
\tau_{\rm s} = (\gtime + \tau_{\rm i2}) - \tau_{\rm i1}
\label{eq:intser}
\end{equation}
where $\tau_{\rm i1}$ and $\tau_{\rm i2}$ represent incubation periods of an infector and
an infectee, respectively, and $\gtime$ represents the generation
interval between the infector and the infectee.
These studies concluded that the serial and generation intervals have the same mean when $\tau_{\rm i1}$ and $\tau_{\rm i2}$ are drawn from the same distributions \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation}.
However, distributions of realized incubation periods, $\tau_{\rm i1}$ and $\tau_{\rm i2}$ will be identical only if we assume that they are intrinsic to individuals (and not dependent on epidemic dynamics at the population-level) ---
something that is generally true of forward but not backward incubation-period distributions.
We refer to the definition \eref{intser} as the intrinsic serial interval (\fref{diagram}A).

@@

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total density of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time \ssymp.
For simplicity, we assume that all infected individuals eventually develop symptoms.
Then, the density of serial intervals between time $\psymp$ and \ssymp,
given that the infectors became infected at time $\pinf\le\psymp$ and
the infectees became infected at time $\sinf\le\ssymp$, depends on the
amount of infection that occurs between time $\pinf$ and $\sinf$ as well as
the density of forward incubation periods between $\pinf$ and
$\psymp$ (realized incubation periods of infectors) and between
$\sinf$ and \ssymp (realized incubation periods of infectees):
\begin{equation}
\total (\psymp,\ssymp | \alpha_1, \alpha_2)=\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\substack{\text{incidence} \\ \text{of} \\ \text{infection}}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{forward incubation} \\ \text{periods } p-\pinf \text{ and forward} \\ \text{generation intervals } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{\idist_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{forward incubation} \\ \text{periods } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases that a primary case infected at time $\pinf$ will generate over the course of their infection \citep{fraser2007estimating}.
We describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

@@

We illustrate changes in forward and backward serial intervals over the course of an epidemic by applying our framework to a specific example of an epidemic model.
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be described as:
\begin{align}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\nonumber\\
i(t) &= \RR(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\label{eq:renewal}
\end{align}
where $\RR(t)$ is the instantaneous reproduction number (i.e., the average number of secondary cases that a primary case infected at time $t$ will generate if conditions at time $t$ remain unchanged \citep{fraser2007estimating}), and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where changes in $\RR(t)$ is negligible \citep{champredon2015intrinsic}).
This model assumes that $g(\tau)$ remains constant through time -- in other words, that epidemic dynamics are driven by changes in transmission rate. 
This assumption may not be well suited to individual-based intervention such as case isolation \citep{fraser2007estimating};
nonetheless, this form has been widely used in the literature and has been successfully applied in modeling the current COVID-19 pandemic \citep{gostic2020practical}.

Here, changes in reproduction number can be modeled as a product of the basic reproduction number $\Ro$, proportion susceptible $S(t)$, and a time-dependent factor $M(t)$ (for example, accounting for nonpharmaceutical interventions and behavioral changes): $\RR(t) = \Ro S(t) M(t)$;
\cite{flaxman2020estimating} used a similar framework to evaluate the impact of nonpharmaceutical interventions on the spread of COVID-19 in 11 countries.
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) = \frac{ \gdist(\tau) S(\psymp + \tau) M(\psymp + \tau) }{\int_0^\infty \gdist(\tau') S(\psymp + \tau') M(\psymp + \tau') \dtau'},
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) = \frac{h(x, \tau) S(\psymp + \tau) M(\psymp + \tau)}{\int_0^\infty \int_0^\infty h(x', \tau') S(\psymp + \tau') M(\psymp + \tau') \dtau' \dx'}.
\end{equation}
We further assume that the forward incubation period distribution does
not vary across cohorts over the course of an epidemic, as it
represents the life history of a disease; we denote it as $\idist$.
Then, we have:
\begin{align}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\nonumber,\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx.
\label{eq:marginal}
\end{align}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) M(\psymp + \tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective (i.e., forward vs. backward).

@@

During the initial phase of an epidemic, the proportion susceptible remains approximately constant ($S(t) \approx S(0)$) and incidence of infection grows exponentially: $i(t) \approx i_0\exp(rt)$.
During this period, the intrinsic generation-interval distribution provides the correct link between the exponential growth rate $r$ and the initial reproduction number $\RR=\Ro S(0)$ based on the Euler-Lotka equation \citep{wallinga2007generation}.
Here, we focus on the estimates of the basic reproduction number \Ro (the value of \RR in a fully susceptible population, $S(t) \approx 1$):
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) g(\tau) \dtau.
\label{eq:Rgen}
\end{equation}
Analogous to the intrinsic generation-interval distribution, 
forward serial-interval distributions describe the renewal process between symptomatic cases.
Therefore, we expect the forward serial-interval distribution during the exponential growth phase --- which we refer to as the \emph{initial} forward serial-interval distribution $f_0$ --- to estimate the same value of \Ro for a given $r$ as the intrinsic generation-interval distribution (note, however, that the forward serial interval is not necessarily positive):
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau.
\label{eq:Rforward}
\end{equation}
Here, the initial forward serial-interval distribution is given by:
\begin{equation}
f_{0}(\tau) = \frac{1}{\phi} \int_{-\infty}^{0} \int_{\pinf}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:initialSI}
\end{equation}
where the normalization constant $\phi$ is determined by the requirement that $\int_{-\infty}^\infty f_{0}(\tau)\,\dtau=1$.
We provide a mathematical proof of this relationship in Supplementary Materials.
Since we do not make any assumptions about the shape of the joint distribution $h$ between incubation periods and the generation intervals, \eref{Rforward} holds in general whether or not there is a presymptomatic transmission period.

We further compare this with the estimate of \Ro based on the intrinsic serial-interval distribution $q(\tau)$:
\begin{equation}
\frac{1}{\Rintrinsic} = \int_{-\infty}^\infty \exp(-r\tau) q(\tau) \mathrm{d} \tau.
\label{eq:Rintrinsic}
\end{equation}
The intrinsic serial-interval distribution $q(\tau)$ does not depend on epidemic dynamics, and is given by:
\begin{equation}
q(\tau) = \frac{1}{\phi_q} \int_{-\infty}^{0} \int_{\pinf}^{\tau} h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:intrinsicSI}
\end{equation}
where the normalization constant $\phi_q$ is determined by the
requirement that $\int_{-\infty}^\infty q(\tau)\,\dtau=1$.
Rather than numerically integrating over closed forms of $g$, $f_0$, and $q$ to estimate $\Ro$, we use simulation-based approaches for simplicity (Supplementary Materials).

The initial forward serial-interval distribution depends on the exponential growth rate $r$.
For a fast-growing epidemic (high $r$), we expect the backward incubation periods to be short (\eref{backexp}), meaning that presymptomatic transmission is less likely to occur. 
Therefore, the initial forward serial-interval distribution will generally have a larger mean than the intrinsic generation- and serial-interval distributions.
However, the exact shape of the initial forward serial-interval distribution depends on the shape of the joint distribution.
For example, the Susceptible-Exposed-Infected-Recovered model, under the additional assumption that the incubation and exposed periods are equivalent (i.e. that onset of symptoms and infectiousness occur simultaneously), provides a special case.
In this case, the forward serial- and generation-intervals follow the same distributions during the exponential growth phase because (i) infected individuals can only transmit after symptom onset and (ii) the time between symptom onset and infection is independent of the incubation period of an infector (see Supplementary Materials).
Everywhere else in this paper, however, we do not assume that the incubation and exposed periods are equivalent.
Instead, we allow for presymptomatic transmission in the model in order to reflect the transmission dynamics of COVID-19.

@@

\section{Supplementary Materials}

\subsection{Deterministic simulation}

>>>>

\usepackage{natbib}
\bibliographystyle{unsrt}
\date{\today}

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic, such as the current pandemic of coronavirus disease 2019 (COVID-19) \cite{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case.
The value in a fully susceptible population --- the ``basic'' reproduction number \Ro --- allows us to predict the extent to which an infection will spread in the population, and the amount of intervention necessary to eliminate it in simple cases \cite{anderson1991infectious}.
Since the reproduction number represents an average \cite{anderson1991infectious,diekmann1990definition}, it fails to capture heterogeneity among individuals or across space.
The reproduction number also fails to provide any information about the time scale of disease transmission.

Estimating the reproduction number \RR is often challenging.
Direct estimates based on observed infections will typically be biased down when some infections cannot be observed.
A common method of estimating \RR near the beginning of an epidemic is based on the population-level exponential growth rate $r$, which can often be estimated robustly from case reports \cite{mills2004transmissibility, ma2014estimating}.
The growth rate $r$ and the reproduction number \RR are linked by the generation-interval distribution \cite{wallinga2007generation}, where the generation interval is defined as the time between when an individual (infector) is infected and when that individual infects another person (infectee) \cite{svensson2007note}.

Since generation intervals measure time between infection events, which can be difficult to observe in practice, generation intervals are often replaced with serial intervals.
The serial interval is defined as the time between when an infector and an infectee develop symptoms \cite{svensson2007note}.
While generation and serial intervals both measure the time scale of disease transmission, they measure fundamentally different quantities. 
In particular, previous studies have noted that, in many contexts, serial intervals are expected to have larger variances than generation intervals but have the same mean in many contexts \cite{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence}.
Serial intervals can in some cases even take negative values in the presence of presymptomatic transmission \cite{he2020temporal}, whereas generation intervals must be positive.

Although these distributions were clearly and distinctly defined over a decade ago \cite{svensson2007note}, 
the need for a better conceptual and theoretical framework for understanding their differences is becoming clearer as the COVID-19 pandemic unfolds.
Researchers continue to base inferences about COVID-19 on both generation and serial intervals without clearly distinguishing between them (e.g., \cite{tempvar,du2020serial,he2020temporal,wu2020nowcasting,zhao2020serial}), and, in some cases, explicitly conflate the definitions of the two intervals (e.g., \cite{anderson2020will,hellewell2020feasibility}).
This confusion is apparent even in standard software for estimating \RR, such as \texttt{EpiEstim}, in which the serial-interval distribution is used to infer time-dependent \RR \cite{thompson2019improved}.
These studies are examples of many---indeed, it is a common practice to use the serial and generation intervals interchangeably.

One source of confusion arises from an apparent discrepancy between the generation-interval and serial-interval viewpoints.
While the epidemic is growing exponentially, the spread of infection can be characterized as a \emph{renewal process} based on previous incidence of infection, the associated generation-interval distribution, and the average infectiousness of an infected individual.
It is well established that this renewal formulation allows us to link the exponential growth rate of an epidemic $r$ with its reproduction number \RR using the generation-interval distribution \cite{wallinga2007generation}.
However, the serial-interval distribution also describes a renewal process --- in this case, the creation of a new \emph{symptomatic} case based on a symptomatic case in the previous generation.
Since both renewal processes, based on either generation- or serial-interval distributions, describe the same underlying exponentially growing system, both should provide the same correct link between the reproduction number \RR and the epidemic growth rate $r$.

In contexts where the serial- and generation-interval distributions differ, current theory has no explanation for how two different distributions could provide identical estimates of \RR from $r$. 
In fact, recent theory suggest that using the serial-interval can underestimate the reproduction number \cite{britton2019estimation, ganyani2020estimating}.
However, these studies rely on \emph{intrinsic} distributions of incubation periods and generation intervals that neglect the population-level dynamics of disease spread.

Here we show that, by correctly defining and calculating the ``forward'' serial-interval distribution (i.e., a distribution of serial intervals from a cohort of infectors that developed symptoms at the same time) that connects symptom onset dates, we can resolve this discrepancy.
These forward intervals are different from the ``intrinsic'' serial intervals that previous studies have relied on \cite{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence, britton2019estimation}.
During an ongoing epidemic, all observed epidemiological delays (e.g., incubation period) between primary (e.g., infection) and secondary (e.g., symptom onset) events are subject to backward biases: when the incidence of primary events is increasing (or decreasing), we are more likely to observe shorter (respectively longer) intervals.
In particular, when we consider forward serial-interval distributions, the incubation periods of the infectors are subject to backward biases because we have to look backward in time from their symptom onset to infection.
Therefore, the realized incubation period distributions of the infector and the infectee can differ dynamically, even if the intrinsic analogues of the same distributions are expected to be equivalent.

@@

At the individual level, we can define the time distribution between a primary and a secondary event that we expect to observe for a single infected individual by averaging across individual characteristics --- we refer to this distribution as the \emph{intrinsic distribution}.
For example, the intrinsic incubation period distribution describes the expected time distribution from infection to symptom onset of an infected individual.
Likewise, the intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by an infected individual.
However, the intrinsic time distributions are not always equivalent to the corresponding realized time distributions at the population level (i.e., the distribution of time between actual primary and secondary events that occur during an epidemic; see \fref{diagram}).
For example, an infectious contact results in infection only if the contacted individual is susceptible (and has not already been infected) ---
this is one mechanism that causes realized generation intervals (time between actual infection events) to differ from from the intrinsic generation intervals (time between infection and infectious contacts) \cite{park2020inferring}.
In this example, the difference between intrinsic and realized time distributions can be attributed to the fact that the fraction of susceptible individuals is itself dynamic.

@@

Several different mechanisms drive the changes in forward and backward delay distributions over time.
Typically, within-individual forward delay distributions are not directly affected by epidemic dynamics.
Some realized forward distributions, like incubation period distributions, are equivalent to their intrinsic distributions and remain invariant at the time scale of an outbreak.
Other realized distributions, like the distribution of time from symptom onset to testing, may change over the course of an epidemic due to changes in public-health policies or individual behavior.
Between-individual forward delay distributions, such as generation- or serial-interval distributions, depend on epidemic dynamics.
For example, forward generation intervals often become shorter as an epidemic progresses due to the dynamical process of susceptible depletion, as well as due to other factors like behavioral change or interventions \cite{kenah2008generation, nishiura2010time, champredon2015intrinsic}: if it is harder to infect later in the course of infection, then proportionally more intervals will be short.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when their infectee becomes symptomatic \cite{svensson2007note}.
Previous studies have often expressed serial intervals $\tau_{\rm s}$ in the form (\fref{diagram}A):
\begin{equation}
\tau_{\rm s} = (\gtime + \tau_{\rm i2}) - \tau_{\rm i1}
\label{eq:intser}
\end{equation}
where $\tau_{\rm i1}$ and $\tau_{\rm i2}$ represent incubation periods of an infector and
an infectee, respectively, and $\gtime$ represents the generation
interval between the infector and the infectee.
These studies concluded that the serial and generation intervals have the same mean when $\tau_{\rm i1}$ and $\tau_{\rm i2}$ are drawn from the same distributions \cite{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation}.
However, distributions of realized incubation periods, $\tau_{\rm i1}$ and $\tau_{\rm i2}$ will be identical only if we assume that they are intrinsic to individuals (and not dependent on epidemic dynamics at the population-level) ---
something that is generally true of forward but not backward incubation-period distributions.
We refer to the definition \eref{intser} as the intrinsic serial interval (\fref{diagram}A).

@@

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total density of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time \ssymp.
For simplicity, we assume that all infected individuals eventually develop symptoms.
Then, the density of serial intervals between time $\psymp$ and \ssymp,
given that the infectors became infected at time $\pinf\le\psymp$ and
the infectees became infected at time $\sinf\le\ssymp$, depends on the
amount of infection that occurs between time $\pinf$ and $\sinf$ as well as
the density of forward incubation periods between $\pinf$ and
$\psymp$ (realized incubation periods of infectors) and between
$\sinf$ and \ssymp (realized incubation periods of infectees):
\begin{equation}
\total (\psymp,\ssymp | \alpha_1, \alpha_2)=\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\substack{\text{incidence} \\ \text{of} \\ \text{infection}}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{forward incubation} \\ \text{periods } p-\pinf \text{ and forward} \\ \text{generation intervals } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{\idist_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{forward incubation} \\ \text{periods } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases that a primary case infected at time $\pinf$ will generate over the course of their infection \cite{fraser2007estimating}.
We describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

@@

We illustrate changes in forward and backward serial intervals over the course of an epidemic by applying our framework to a specific example of an epidemic model.
We model disease spread with a renewal-equation model \cite{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be described as:
\begin{align}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\nonumber\\
i(t) &= \RR(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\label{eq:renewal}
\end{align}
where $\RR(t)$ is the instantaneous reproduction number (i.e., the average number of secondary cases that a primary case infected at time $t$ will generate if conditions at time $t$ remain unchanged \cite{fraser2007estimating}), and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where changes in $\RR(t)$ is negligible \cite{champredon2015intrinsic}).
This model assumes that $g(\tau)$ remains constant through time -- in other words, that epidemic dynamics are driven by changes in transmission rate. 
This assumption may not be well suited to individual-based intervention such as case isolation \cite{fraser2007estimating};
nonetheless, this form has been widely used in the literature and has been successfully applied in modeling the current COVID-19 pandemic \cite{gostic2020practical}.

Here, changes in reproduction number can be modeled as a product of the basic reproduction number $\Ro$, proportion susceptible $S(t)$, and a time-dependent factor $M(t)$ (for example, accounting for nonpharmaceutical interventions and behavioral changes): $\RR(t) = \Ro S(t) M(t)$;
\cite{flaxman2020estimating} used a similar framework to evaluate the impact of nonpharmaceutical interventions on the spread of COVID-19 in 11 countries.
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \cite{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) = \frac{ \gdist(\tau) S(\psymp + \tau) M(\psymp + \tau) }{\int_0^\infty \gdist(\tau') S(\psymp + \tau') M(\psymp + \tau') \dtau'},
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) = \frac{h(x, \tau) S(\psymp + \tau) M(\psymp + \tau)}{\int_0^\infty \int_0^\infty h(x', \tau') S(\psymp + \tau') M(\psymp + \tau') \dtau' \dx'}.
\end{equation}
We further assume that the forward incubation period distribution does
not vary across cohorts over the course of an epidemic, as it
represents the life history of a disease; we denote it as $\idist$.
Then, we have:
\begin{align}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\nonumber,\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx.
\label{eq:marginal}
\end{align}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) M(\psymp + \tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective (i.e., forward vs. backward).

@@

During the initial phase of an epidemic, the proportion susceptible remains approximately constant ($S(t) \approx S(0)$) and incidence of infection grows exponentially: $i(t) \approx i_0\exp(rt)$.
During this period, the intrinsic generation-interval distribution provides the correct link between the exponential growth rate $r$ and the initial reproduction number $\RR=\Ro S(0)$ based on the Euler-Lotka equation \cite{wallinga2007generation}.
Here, we focus on the estimates of the basic reproduction number \Ro (the value of \RR in a fully susceptible population, $S(t) \approx 1$):
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) g(\tau) \dtau.
\label{eq:Rgen}
\end{equation}
Analogous to the intrinsic generation-interval distribution, 
forward serial-interval distributions describe the renewal process between symptomatic cases.
Therefore, we expect the forward serial-interval distribution during the exponential growth phase --- which we refer to as the \emph{initial} forward serial-interval distribution $f_0$ --- to estimate the same value of \Ro for a given $r$ as the intrinsic generation-interval distribution (note, however, that the forward serial interval is not necessarily positive):
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau.
\label{eq:Rforward}
\end{equation}
Here, the initial forward serial-interval distribution is given by:
\begin{equation}
f_{0}(\tau) = \frac{1}{\phi} \int_{-\infty}^{0} \int_{\pinf}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:initialSI}
\end{equation}
where the normalization constant $\phi$ is determined by the requirement that $\int_{-\infty}^\infty f_{0}(\tau)\,\dtau=1$.
We provide a mathematical proof of this relationship in SI Appendix S3.
Since we do not make any assumptions about the shape of the joint distribution $h$ between incubation periods and the generation intervals, \eref{Rforward} holds in general whether or not there is a presymptomatic transmission period.

We further compare this with the estimate of \Ro based on the intrinsic serial-interval distribution $q(\tau)$:
\begin{equation}
\frac{1}{\Rintrinsic} = \int_{-\infty}^\infty \exp(-r\tau) q(\tau) \mathrm{d} \tau.
\label{eq:Rintrinsic}
\end{equation}
The intrinsic serial-interval distribution $q(\tau)$ does not depend on epidemic dynamics, and is given by:
\begin{equation}
q(\tau) = \frac{1}{\phi_q} \int_{-\infty}^{0} \int_{\pinf}^{\tau} h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf,
\label{eq:intrinsicSI}
\end{equation}
where the normalization constant $\phi_q$ is determined by the
requirement that $\int_{-\infty}^\infty q(\tau)\,\dtau=1$.
Rather than numerically integrating over closed forms of $g$, $f_0$, and $q$ to estimate $\Ro$, we use simulation-based approaches for simplicity (see SI Appendix S4).

The initial forward serial-interval distribution depends on the exponential growth rate $r$.
For a fast-growing epidemic (high $r$), we expect the backward incubation periods to be short (\eref{backexp}), meaning that presymptomatic transmission is less likely to occur. 
Therefore, the initial forward serial-interval distribution will generally have a larger mean than the intrinsic generation- and serial-interval distributions.
However, the exact shape of the initial forward serial-interval distribution depends on the shape of the joint distribution.
For example, the Susceptible-Exposed-Infected-Recovered model, under the additional assumption that the incubation and exposed periods are equivalent (i.e. that onset of symptoms and infectiousness occur simultaneously), provides a special case.
In this case, the forward serial- and generation-intervals follow the same distributions during the exponential growth phase because (i) infected individuals can only transmit after symptom onset and (ii) the time between symptom onset and infection is independent of the incubation period of an infector (see SI Appendix S5).
Everywhere else in this paper, however, we do not assume that the incubation and exposed periods are equivalent.
Instead, we allow for presymptomatic transmission in the model in order to reflect the transmission dynamics of COVID-19.

@@

\section*{Competing interests}

We declare no competing interests.
